---
title: "Data Wrangling with dplyr"
author: "Karen Dworsky"
date: "2025-09-08"
format: html
execute:
  cache: true
---

## Setup

```{r}
library(tidyverse)
library(palmerpenguins)
data("penguins")
```

## Question 1

Fixing common errors: `c()` for multiple columns, correcting spelling, specifying the data frame name.

```{r}
penguins[1:5, c("species", "island")]
```

## Question 2

The species with the largest average flipper length is Gentoo at 217 mm, followed by Chinstrap at 196 mm, and smallest Adelie at 190 mm.

```{r}
penguins %>%
  group_by(species) %>%
  summarise(flipper_length_avg = mean(flipper_length_mm, na.rm = TRUE))
```

## Question 3

Adelie penguins are on all three islands, whereas Chinstrap and Gentoo are each on one island.

```{r}
penguins %>%
  group_by(species, island) %>%
  summarise(n_observations = n())
```

## Question 4

Reorder columns to put year first.

```{r}
penguins %>%
  select(year, everything())
```

## Question 5

Create a size group column based on whether each penguin is above or below the mean body mass.

```{r}
penguins_fixed <- penguins %>%
  mutate(size_group = if_else(
    body_mass_g > mean(penguins$body_mass_g, na.rm = TRUE),
    paste0(species, " large"),
    paste0(species, " small")
  ))
penguins_fixed %>% select(size_group)
```

## Question 6

Hypothesis: larger penguins cluster by island, suggesting one island has better food resources.

```{r}
ggplot(penguins) +
  geom_point(mapping = aes(x = body_mass_g, y = flipper_length_mm, color = island))
```

Biscoe Island has both the largest overall body masses. Checking the range:

```{r}
penguins_clean <- penguins %>% filter(!is.na(body_mass_g))

island_stats <- penguins_clean %>%
  group_by(island) %>%
  summarise(
    n_penguins = n(),
    min_body_mass = min(body_mass_g),
    max_body_mass = max(body_mass_g)
  )

island_stats
```
