---
title: "Principles of Data Visualization"
author: "Karen Dworsky"
date: "2025-09-13"
format: html
execute:
  cache: true
---

## Setup

```{r}
library(ggplot2)
library(palmerpenguins)
library(dplyr)

data(penguins)
penguins_clean <- penguins %>%
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm),
         !is.na(flipper_length_mm), !is.na(body_mass_g))
```

## The Good Plot

A well-designed plot: clear title and axis labels, colorblind-friendly palette, appropriate point size, regression lines per species, minimal theme.

```{r}
good_plot <- penguins_clean %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_manual(
    values = c("Adelie" = "#FF6B35", "Chinstrap" = "#004E89", "Gentoo" = "#009639"),
    name = "Species"
  ) +
  labs(
    title = "Relationship Between Flipper Length and Body Mass in Palmer Penguins",
    subtitle = "Larger penguins tend to have longer flippers across all species",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    caption = "Data source: Palmer Penguins dataset"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 11, color = "gray30", hjust = 0),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(size = 9, color = "gray50")
  )

print(good_plot)
```

## The Bad Plot

A deliberately poor plot demonstrating common mistakes: misleading title, clashing colors (not colorblind-friendly), angled axis labels, arbitrary axis limits that cut data, garish background colors.

```{r}
bad_plot <- penguins_clean %>%
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = sex)) +
  geom_point(size = 8, alpha = 1) +
  scale_color_manual(
    values = c("female" = "red", "male" = "green"),
    name = "Gender of the Penguin Bird"
  ) +
  labs(
    title = "The best graph in the world",
    x = "The length of a panguin bill in millimeters unit",
    y = "mm bill depth of penguin"
  ) +
  xlim(30, 65) +
  ylim(12, 25) +
  theme(
    plot.title = element_text(size = 20, face = "bold", color = "purple"),
    axis.title.x = element_text(size = 14, angle = 45, color = "blue"),
    axis.title.y = element_text(size = 14, angle = 45, color = "orange"),
    axis.text.x = element_text(angle = 90, size = 6),
    axis.text.y = element_text(angle = 45, size = 6),
    legend.position = "top",
    legend.title = element_text(size = 8),
    panel.background = element_rect(fill = "yellow"),
    plot.background = element_rect(fill = "lightblue"),
    panel.grid.major = element_line(color = "darkgreen", size = 2),
    panel.grid.minor = element_line(color = "pink", size = 1)
  )

print(bad_plot)
```
